Le titre

### 2022-08-08 Manuscrit Thèse/chapter 3/Chapter 3.tex
t

\section{Experimental observation of NV-NV cross-relaxation}
Before we discuss the theoretical complications related to NV-NV cross-relaxations (CR), let us first show the unambiguous experimental proof of the presence of NV-NV CR.
\subsection{NV-NV CR between nonequivalent NV centers}
\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{Figures/NV-NV_non_equivalent}
\caption{label. Taken from \citep{armstrong2010nv}}
\label{non equivalent NV-NV}
\end{figure}
NV-NV CR was first observed more than thirty years ago \citep{holliday1989optical, van1989cross}. The first observations were between non-equivalent NV centers, meaning that the two NV centers involved in the dipole-dipole coupling were not polarized equally.

@@

\section{Validation of the fluctuator model}
We will now show experimental results validating the fluctuator model.
\subsection{The stretched exponential lifetime}
\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{Figures/Stretch_vs_pas_stretch}
\label{stretch_or_not_stretch}
\caption{T1 measurement following the protocol described in [REF]. The best exponential and stretched exponential fits are given each time. a) Sample ADM-15-2 with B=0. The optimal $T_1$ values are $T_1^{\rm stretch}=150\ \mu \rm s$ and $T_1^{\rm exp}=410\ \mu \rm s$. b) Sample CVD-pink with B$\neq$0 and a single class probed. The optimal $T_1$ values are $T_1^{\rm stretch}=1.9\ \rm ms$ and $T_1^{\rm exp}=4.1\ \rm ms$}
\end{figure}
$T_1$ measurements with dense ensemble of NV centers and when many classes are resonant with each other (typically when B=0) indeed tend to show stretched exponential profiles.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{Figures/various_fit_formulae}
\label{various_fit_formulae}
\caption{Different fitting procedure on two $T_1$ measurements done on sample ADM-150-1 for $B=0$ and $B\neq0$. \\ a) Exponential fit $S(\tau)=\exp (-\tau/T_1^{\rm ph})$. \\ b) Stretched exponential fit $S(\tau)=\exp (-\sqrt{\tau/T_1^{\rm dd}})$. \\ c) Bi-exponential fit $S(\tau)=\exp (-\tau/T_1^{\rm ph} -\sqrt{\tau/T_1^{\rm dd}})$. \\ d) Bi-exponential fit with fixed $T_1^{\rm ph}=5\ \rm ms$: $S(\tau)=\exp (-\tau/5 -\sqrt{\tau/T_1^{\rm dd}})$.}
\end{figure}

\begin{table}[htbp]
\centering
\caption{Fitting parameters on Fig. \ref{various_fit_formulae}}
\begin{tabular}{c|cc|cc}
\toprule
Figure &  \multicolumn{2}{c}{$B=0$} & \multicolumn{2}{c}{$B\neq0$}\\
\midrule
{} &  $T_1^{\rm ph}$ (ms)& $T_1^{\rm dd}$ (ms)&  $T_1^{\rm ph}$ (ms)& $T_1^{\rm dd}$(ms)\\
Fig. a) & 0.95 & * & 2.64 & * \\
Fig. b) & * & 0.32 & * & 1.03  \\
Fig. c) & 6.75 & 0.44 & 4.22 & 7.19 \\
Fig. d) & \textbf{5} & 0.51 & \textbf{5} & 4.65 \\
\bottomrule
\end{tabular}

Bold characters indicate parameters that were arbitrarily fixed.
  \label{fitting table}
\end{table}

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{Figures/betas}
\label{betas}
\caption{a) Same two measurements as Fig. \ref{various_fit_formulae} with the fitting formula $S(\tau)=\exp ((-\tau/T_1)^{\beta})$. The fitting parameters are $\beta=0.58$ and $T_1=0.46\ \rm ms$ for $B=0$, and $\beta=0.80$ and $T_1=2.15\ \rm ms$ for $B\neq0$. b) Optimal $\beta$ parameter found for each $T_1$ measurement as a function of the external magnetic field (still on sample ADM-150-1).}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{Figures/fluctuator linewidth}
\label{fluct linewidth}
\caption{Measurement of the fluctuators linewidth using the same setup as in Fig. \ref{equivalent NV-NV}. a) Simulation of the $\ket{0} \to \ket{-1}$ transition frequency for the four classes of NV centers and actual frequencies of the two central classes from ODMR measurements. b) Measurement of $T_1^{\rm dd}$ as a function of the splitting $\Delta \nu$ between the two central classes, fitted with a Lorentzian of half-width $\sigma=8.78\ \rm MHz$. The green dashed line correspond to the ODMR spectrum of a single class of NV centers.}
\end{figure}

@@

Fig. \ref{fluct linewidth} shows such

...Note that only the two classes ... were measured accurately since ... The central region ... overlap ... linear regression
\subsection{Parallel with P-doped Si}
"Electron Spin-Lattice Relaxation in Phosphorus-Doped Silicon"

>>>>

\section{Experimental observation of NV-NV cross-relaxation}
Before we discuss the theoretical complications related to NV-NV cross-relaxations (CR), let us first show the unambiguous experimental proof of the presence of NV-NV CR with dense NV ensembles ([NV]$\gtrsim$ 1 ppm).
\subsection{NV-NV CR between nonequivalent NV centers}
\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{Figures/NV-NV_non_equivalent}
\caption{label. Taken from \citep{armstrong2010nv}}
\label{non equivalent NV-NV}
\end{figure}
NV-NV CR was first observed more than thirty years ago \citep{holliday1989optical, van1989cross}. The first observations were between non-equivalent NV centers, meaning that the two NV centers involved in the dipole-dipole coupling were not polarized equally.

@@

\section{Experimental investigation of the fluctuator model}
We will now show experimental results related to the predictions of the fluctuator model.
\subsection{The stretched exponential lifetime}
\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{Figures/Stretch_vs_pas_stretch}
\caption{T1 measurement following the protocol described in [REF]. The best exponential and stretched exponential fits are given each time. a) Sample ADM-15-2 with B=0. The optimal $T_1$ values are $T_1^{\rm stretch}=150\ \mu \rm s$ and $T_1^{\rm exp}=410\ \mu \rm s$. b) Sample CVD-pink with B$\neq$0 and a single class probed. The optimal $T_1$ values are $T_1^{\rm stretch}=1.9\ \rm ms$ and $T_1^{\rm exp}=4.1\ \rm ms$}
\label{stretch_or_not_stretch}
\end{figure}
$T_1$ measurements with dense ensemble of NV centers and when many classes are resonant with each other (typically when B=0) indeed tend to show stretched exponential profiles.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{Figures/various_fit_formulae}
\caption{Different fitting procedure on two $T_1$ measurements done on sample ADM-150-1 for $B=0$ and $B\neq0$. \\ a) Exponential fit $S(\tau)=\exp (-\tau/T_1^{\rm ph})$. \\ b) Stretched exponential fit $S(\tau)=\exp (-\sqrt{\tau/T_1^{\rm dd}})$. \\ c) Bi-exponential fit $S(\tau)=\exp (-\tau/T_1^{\rm ph} -\sqrt{\tau/T_1^{\rm dd}})$. \\ d) Bi-exponential fit with fixed $T_1^{\rm ph}=5\ \rm ms$: $S(\tau)=\exp (-\tau/5 -\sqrt{\tau/T_1^{\rm dd}})$.}
\label{various_fit_formulae}
\end{figure}

\begin{table}[htbp]
\centering
\caption{Fitting parameters for Fig.
 \ref{various_fit_formulae}}
 \label{fitting table}
\begin{tabular}{c|cc|cc}
\toprule
Figure &  \multicolumn{2}{c}{$B=0$} & \multicolumn{2}{c}{$B\neq0$}\\
\midrule
{} &  $T_1^{\rm ph}$ (ms)& $T_1^{\rm dd}$ (ms)&  $T_1^{\rm ph}$ (ms)& $T_1^{\rm dd}$(ms)\\
Fig. a) & 0.95 & * & 2.64 & * \\
Fig. b) & * & 0.32 & * & 1.03  \\
Fig. c) & 6.75 & 0.44 & 4.22 & 7.19 \\
Fig. d) & \textbf{5} & 0.51 & \textbf{5} & 4.65 \\
\bottomrule
\end{tabular}

Bold characters indicate parameters that were arbitrarily fixed.

\end{table}

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{Figures/betas}
\caption{a) Same two measurements as Fig. \ref{various_fit_formulae} with the fitting formula $S(\tau)=\exp ((-\tau/T_1)^{\beta})$. The fitting parameters are $\beta=0.58$ and $T_1=0.46\ \rm ms$ for $B=0$, and $\beta=0.80$ and $T_1=2.15\ \rm ms$ for $B\neq0$. b) Optimal $\beta$ parameter found for each $T_1$ measurement as a function of the external magnetic field (still on sample ADM-150-1).}
\label{betas}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{Figures/fluctuator linewidth}
\caption{Measurement of the fluctuators linewidth using the same setup as in Fig. \ref{equivalent NV-NV}. a) Simulation of the $\ket{0} \to \ket{-1}$ transition frequency for the four classes of NV centers and actual frequencies of the two central classes from ODMR measurements. b) Measurement of $T_1^{\rm dd}$ as a function of the splitting $\Delta \nu$ between the two central classes, fitted with a Lorentzian of half-width $\sigma=8.78\ \rm MHz$. The green dashed line correspond to the ODMR spectrum of a single class of NV centers, and the red dashed one to the same line scaled up by a factor $\sqrt{2}$}
\label{fluct linewidth}
\end{figure}

@@

Fig. \ref{fluct linewidth} shows an experiment similar to the one presented in Fig. \ref{equivalent NV-NV}. A variable magnetic $B_{\rm scan}$ in addition to an offset magnetic field $B_{\rm DC} \sim 100\ \rm G$ are used in order to create a crossing between two classes of NV centers, but this time measuring the spin decay rate instead of the PL.

Fig. \ref{fluct linewidth}-a) shows the simulated transitions of the four classes of NV centers as a function of the magnetic field. Because the $T_1$ measurement protocol described in [REF] requires a resonant microwave pulse with the NV probed, we specifically record ODMR spectra of the two central lines to get a precise measurement of the Larmor frequency. In the region where the two ODMR lines overlap we have to use a linear regression to get the Larmor frequency of each class.

Fig. \ref{fluct linewidth}-b) shows the dipole-dipole induced relaxation rate $T_1^{\rm dd}$ as a function of the detuning $\Delta \nu$ between the two central classes. For each magnetic field value, a $T_1$ profile was recorded and fitted following the procedure described in Fig. \ref{various_fit_formulae}-d) where we again fixed $T_1^{\rm ph}=5\ \rm ms$. We also show the ODMR spectrum of a single NV class for comparison.

We can clearly see that the increase in $T_1^{\rm dd}$ is much broader than the ODMR line of the NV centers, whereas in a standard CR process, we expect the decay rate to be proportional to the spectral overlap of the two classes \citep{hall2016detection}. If we assume that both NV classes spectral response is a gaussian of same width $\sigma$, then the spectral overlap between the two classes is also a gaussian of width $\sqrt{2} \sigma$. Fig. \ref{fluct linewidth}-b) also shows the ODMR line scaled by a factor $\sqrt{2}$, which is still much narrower than the $1/T_1^{\rm dd}$ line.

A potential explanation for the broadening of the $1/T_1^{\rm dd}$ line could be the interaction strength between the NV centers, but for [NV]$\sim$5 ppm, $\expval{\mathcal{H}_{\rm dd}} \sim 46\ \rm kHz$ which cannot explain the $\sim 4\ \rm MHz$ broadening that we observe.

\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth]{Figures/deconvolution}
\caption{Blue curve $1/T_1^{\rm dd}$ data presented in Fig. \ref{fluct linewidth}. Red curve: deconvolution of the blue curve the  by the ODMR line shown on the same figure. Green curve: deconvolution of the red curve the  by the ODMR line. All three curves are fitted with Lorentzian of respective half widths 8.78, 7.49 and 6.38}
\label{deconvolution}
\end{figure}

The fluctuator model provides another explanation: the fluctuator spectral response is broaden by its very short lifetime. This also explains why the $1/T_1^{\rm dd}$ line is well matched by a Lorenzian, while the ODMR lines are closer to Gaussians in our sample.

We can now extract the fluctuator lifetime from the data in Fig. \ref{fluct linewidth} by taking out the contribution of $T_2^*$ on both the NV and the fluctuator spectral response.

The first way to do this is to follow the instructions in \citep{hall2016detection} and deconvolve the $1/T_1^{\rm dd}$ data by the ODMR spectrum. The result of the deconvolution is shown in Fig. \ref{deconvolution}: the data is deconvolved once to get the spectral response of the fluctuator (red curve), and a second time to get only the lifetime contribution of the fluctuator (red curve). Fitting with Lorentzians, we find a final value $\Gamma_2 =\frac{\gamma_f}{2} = (2\pi) 6.38\ \rm MHz$ which corresponds to a fluctuator lifetime $T_1^f=\frac{1}{\gamma_f}=12.5\ \rm ns$.

Another, simpler alternative is to consider that the $1/T_1^{\rm dd}$ profile is a Voigt profile given by the convolution of two Gaussians of half width at half maximum (HWHM) $f_G=\sqrt{2 \ln (2)}\sigma$ and a Lorentzian of HWHM $f_L$:
\begin{align*}
\Gamma_1^{\rm dd}&\propto \mathcal{G}(f_G)*(\mathcal{L}(f_L)*\mathcal{G}(f_G)) \\
&\propto \mathcal{L}(f_L)*\mathcal{G}(\sqrt{2} f_G),
\end{align*}
where $\mathcal{G}(f_G)$ and $\mathcal{L}(f_L)$ denote Gaussian and Lorentzian functions of HWHM $f_G$ and $f_L$ respectively.
Then We can use the formula of a pseudo-Voigt profile which gives an approximation of the Voigt profile's HWHM $f$ as a function of $f_G$ and $f_L$ \citep{wiki:VoigtProfile}:
\begin{equation}
f = [f_G^5 + 2.69269 f_G^4 f_L + 2.42843 f_G^3 f_L^2 + 4.47163 f_G^2 f_L^3 + 0.07842 f_G f_L^4 + f_L^5]^{1/5}.
\end{equation}

With $f=8.78\ \rm MHz$ and $\sqrt{2}f_G=4.33\ \rm MHz$, I find $f_L=6.54\ \rm MHz$ for a final lifetime value $T_1^f = 12.2\ \rm ns$, a value consistent with the one found with the deconvolution method.

Doing similar calculations, \citep{choi2017depolarization} found $\gamma_f=1/T_1^f=3.3\ \rm MHz$, although given their experimental values - full width Half maximum (FWHM) of 25 MHz for the lifetime profile and 9.3 MHz for the ODMR line - I obtain using my method $\gamma_f= 15.7\ \rm MHz$ and $T_1^f=10.1\ \rm ns$. The main difference comes from the fact that they considered the spectral half-width of the fluctuators to be $2\gamma_f$ whereas I consider it to be $\gamma_f/2$.

\section{Microscopic explanation of the fluctuators}
\subsection{Parallel with P-doped Si}
A similar 
"Electron Spin-Lattice Relaxation in Phosphorus-Doped Silicon"
